Query Optimization for Dynamic Imputation

### 2017-03-01 paper/algorithm.tex
paper: polishing changes

\subsection{Imputation operators}
\label{sec:operators}
We introduce two new relational operators to perform imputation: \textit{Impute} ($\mu$) and
\textit{Drop} ($\delta$). Each operator takes arguments $(C, R)$ where $C$ is a set of
attributes and $R$ is a relation. \textit{Impute} uses a machine learning algorithm to
replace all \nullv{} values with non-null values for attributes in $C$ in the relation $R$ (discussed in detail in~\Cref{sec:imputation}).
In our experiments, we instantiate \textit{Impute} with the CE-CART\cite{burgette2010multiple} algorithm, but we highlight that
our system is agnostic to this choice.
\textit{Drop} simply removes from $R$ all tuples which have a \nullv{} value for some attribute in $C$.
Both operators guarantee that the resulting relation will contain no \nullv{} values for
attributes in $C$.

@@

Each imputation operator incurs a heuristic loss penalty. We define \textsc{Loss} as the average over the individual penalties incurred
by $\mu,\delta$ operators. Our recursive definition of $L$ accumulates these penalties in a multi-set, as the order of the penalties is irrelevant,
and averages them at the root of the plan. For example, $\delta$ contributes a loss of 1 (i.e. we lose all information  that dropped tuples
might have encoded). Meanwhile, imputing \nullv{} fields with $\mu$ incurs a loss penalty $p \in (0, 1]$. In this case, the penalty is
inversely proportional to the square root of the number of cells available to train the imputation algorithm, which is defined as
as the product of the number of attributes and the number of tuples. Intuitively, imputation
accuracy \textit{increases} when more complete attributes and complete tuples are available;
correspondingly, the information lost \textit{decreases} with more values to work
with. The square root captures the intuition that the benefits provided by more training observations has diminishing returns.

@@

Scan costs are estimated based on the number of tuples, 
page size, and an unit-less I/O cost per page. Join costs are estimated as a function
of the two relations joined, any blocking operations, and repeated fetching of tuples (if required).
Filtering, projecting, and aggregating are all done in memory and have low computational overhead
so we assume negligible time costs for those operations. We extend these heuristics to 
account for the new operators: \textit{Drop} and \textit{Impute}. \textit{Drop} is a special case of a sequential scan, and its time
complexity can be expressed as a function of the number of heap pages read and the I/O cost
per page.

@@

In general, evaluating the time complexity for different \textit{Impute} operators is
complicated, as the properties of
different algorithms can vary significantly, the underlying I/O cost
is not easily extracted (especially in the case that the entire set of tuples does not
fit in memory), and the CPU cost dominates (in contrast to the other operators, which may
not even consider CPU cost explicitly). For example, \cite{martin1995time} find that the
time complexity of the \textit{build tree} algorithm for one commonly-used class of
decision trees is a function of the number of classes, several overhead constants, the
parameterization of the partition and heuristic functions, and the \textit{arity} (the
number of subsets considered for each split). Indeed, the \textit{build tree} phase often
does not even dominate computation, as post-processing steps like pruning, which are vital
in achieving good performance, can have cost exponential in the height of the tree. In
practice, we find that a simple parameterization can be acceptable (i.e. yield intelligent query
plans).

>>>>

\subsection{Imputation operators}
\label{sec:operators}
We introduce two new relational operators to perform imputation: \textit{Impute} ($\mu$) and
\textit{Drop} ($\delta$). Each operator takes arguments $(C, R)$ where $C$ is a set of
attributes and $R$ is a relation. \textit{Impute} uses a machine learning algorithm to
replace all \nullv{} values with non-null values for attributes in $C$ in the relation $R$ (discussed in detail in~\Cref{sec:imputation}).
In our experiments, we instantiate \textit{Impute} with the CE-CART~\cite{burgette2010multiple} algorithm, but we highlight that
our system is agnostic to this choice.
\textit{Drop} simply removes from $R$ all tuples which have a \nullv{} value for some attribute in $C$.
Both operators guarantee that the resulting relation will contain no \nullv{} values for
attributes in $C$.

@@

Each imputation operator incurs a heuristic loss penalty. We define \textsc{Loss} as the average over the individual penalties incurred
by $\mu,\delta$ operators. Our recursive definition of $L$ accumulates these penalties in a multi-set, as the order of the penalties is irrelevant,
and averages them at the root of the plan. For example, $\delta$ contributes a loss of 1 (i.e. we lose all information  that dropped tuples
might have encoded). Meanwhile, imputing \nullv{} fields with $\mu$ incurs a loss penalty $p \in (0, 1]$. In this case, the penalty is
inversely proportional to the square root of the number of cells available to train the imputation algorithm, which is defined as
the product of the number of attributes and the number of tuples. Intuitively, imputation
accuracy \textit{increases} when more complete attributes and complete tuples are available;
correspondingly, the information lost \textit{decreases} with more values to work
with. The square root captures the intuition that the benefits provided by more training observations has diminishing returns.

@@

Scan costs are estimated based on the number of tuples, 
page size, and a unit-less I/O cost per page. Join costs are estimated as a function
of the two relations joined, any blocking operations, and repeated fetching of tuples (if required).
Filtering, projecting, and aggregating are all done in memory and have low computational overhead
so we assume negligible time costs for those operations. We extend these heuristics to 
account for the new operators: \textit{Drop} and \textit{Impute}. \textit{Drop} is a special case of a sequential scan, and its time
complexity can be expressed as a function of the number of heap pages read and the I/O cost
per page.

@@

In general, evaluating the time complexity for different \textit{Impute} operators is
complicated, as the properties of
different algorithms can vary significantly, the underlying I/O cost
is not easily extracted (especially in the case that the entire set of tuples does not
fit in memory), and the CPU cost dominates (in contrast to the other operators, which may
not even consider CPU cost explicitly). For example,~\cite{martin1995time} find that the
time complexity of the \textit{build tree} algorithm for one commonly-used class of
decision trees is a function of the number of classes, several overhead constants, the
parameterization of the partition and heuristic functions, and the \textit{arity} (the
number of subsets considered for each split). Indeed, the \textit{build tree} phase often
does not even dominate computation, as post-processing steps like pruning, which are vital
in achieving good performance, can have cost exponential in the height of the tree. In
practice, we find that a simple parameterization can be acceptable (i.e. yield intelligent query
plans).

### 2017-03-01 paper/conclusion.tex
paper: polishing changes

\textbf{Imputation confidence:} Obtaining a confidence measure for query results
produced when missing values are imputed with a machine learning model is
another possible improvement to the system. In most cases, the data imputed
is unlikely to follow a standard distribution, which implies that any confidence
estimates will result from iterative approaches such as cross-validation\cite{kohavi1995study}.
In contrast to standard confidence calculations, there is the added complexity
of imputation taking place at various points in the plan and the composition of results
through standard query operations.

>>>>

\textbf{Imputation confidence:} Obtaining a confidence measure for query results
produced when missing values are imputed with a machine learning model is
another possible improvement to the system. In most cases, the data imputed
is unlikely to follow a standard distribution, which implies that any confidence
estimates will result from iterative approaches such as cross-validation~\cite{kohavi1995study}.
In contrast to standard confidence calculations, there is the added complexity
of imputation taking place at various points in the plan and the composition of results
through standard query operations.

### 2017-03-01 paper/experiments.tex
paper: add SMAPE citation
paper: polishing changes

\subsection{Data sets}\label{subsec:datasets}
We collected three data sets for our experiments.
For all data sets, we selected a subset of the original attributes.
We also transformed all data values to an integer representation by enumerating strings and transforming floating point values into an appropriate range.

@@

Our first set of queries is on the CDC data (\Cref{fig:queries-cdc}).
\Cref{q1} calculates calculate the average cuff size for users based on their income data.
\Cref{q2} compares creatine levels for individuals with low, medium, and high incomes spectrum and above a certain weight.
\Cref{q3} extracts the average blood lead levels for children under the age of 6 years.
\Cref{q4} calculates the average systolic blood pressure, by gender, for subjects with a body mass index indicating obesity.

@@

\Cref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error (SMAPE)\todo{cite SMAPE} for
\ProjectName{}'s query results when compared to running imputation on the base tables and
executing the query on this cleaned base data \cite{tofallis2015better}.  Each query was
separately run 200 times in both settings and then results from the two approaches were
paired up and compared tuple-wise.  We average tuple-wise absolute percentage deviations
within each iteration of a query, and we report this value averaged over all iterations.  We
can see that optimizing for quality indeed reduces the SMAPE of query results.  In general, the SMAPE relative 
to the base-imputation approach are low in all cases --- between
\lowsmapealphazero{} and \highsmapealphaoneexacs{} percent ---, indicating that on-the-fly
imputation produces similar results to imputation at the base tables.

>>>>

\subsection{Data sets}\label{subsec:datasets}
We collected three data sets for our experiments.
For all data sets, we selected a subset of the original attributes.
We also transformed all data values to an integer representation by enumerating strings and transforming floating-point values into an appropriate range.

@@

Our first set of queries is on the CDC data (\Cref{fig:queries-cdc}).
\Cref{q1} calculates calculate the average cuff size for users based on their income data.
\Cref{q2} compares creatine levels for individuals with low, medium, and high incomes and above a certain weight.
\Cref{q3} extracts the average blood lead levels for children under the age of 6 years.
\Cref{q4} calculates the average systolic blood pressure, by gender, for subjects with a body mass index indicating obesity.

@@

\Cref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error (SMAPE)~\cite{Makridakis2000451} for
\ProjectName{}'s query results when compared to running imputation on the base tables and
executing the query on this cleaned base data~\cite{tofallis2015better}.  Each query was
separately run 200 times in both settings and then results from the two approaches were
paired up and compared tuple-wise.  We average tuple-wise absolute percentage deviations
within each iteration of a query, and we report this value averaged over all iterations.  We
can see that optimizing for quality indeed reduces the SMAPE of query results.  In general, the SMAPE relative 
to the base-imputation approach are low in all cases --- between
\lowsmapealphazero{} and \highsmapealphaoneexacs{} percent ---, indicating that on-the-fly
imputation produces similar results to imputation at the base tables.

### 2017-03-01 paper/introduction.tex
paper: polishing changes

A simpler approach might drop all rows with missing values, but not only can this introduce bias into the results, but this may result in discarding all of the data.
In contrast to existing systems~\cite{burgette2010multiple,akande2015empirical}, \ProjectName{} avoids imputing over the entire data set, as done with traditional methods. Specifically, \ProjectName{}
carefully plans the placement  imputation / row drop operations inside the query plan, resulting in a 
significant speedup in query execution while minimizing the number of dropped rows.  Also unlike previous work, the focus on of our work is not on the imputation algorithms themselves (we can employ almost any such algorithm), but rather on the  query planning involved in placement of imputation operations
in an optimal way into query plans.   Specifically, our optimization algorithms generate the Pareto-optimal frontier of plans that trade imputation cost for result quality.

@@

\subsection{Contributions}
\ProjectName{} leverages a number of contributions to minimize imputation time while producing comparable results to traditional approaches,
increasing productivity for users who frequently encounter missing values in their data. These contributions include:
\begin{itemize}
\item \textbf{Relational algebra extended with imputation:}
  We extend the standard relational algebra with two new operators to represent imputation operations: impute ($\mu$) and drop ($\delta$).
  Impute operation fills in missing data values using an iterative regression model (CE-CART)~\cite{burgette2010multiple}.
  Drop operation simply drops tuples which contain \nullv{} values.
\item \textbf{Model of imputation quality and cost:}
  We extend the traditional cost model for query plans to incorporate a measure of the quality of the imputations performed.
  We use the cost model to abstract over the imputation algorithm used.
  To add an additional imputation technique, it is sufficient to characterize it with two functions: one to describe its running time and one to describe the quality of its results.
\item \textbf{Query planning with imputation:}
  We present the first query planning algorithm to jointly optimize for running time and the quality of the imputed results.
  It does so by maintaining multiple sets of Pareto optimal plans according to the cost model.
  By deferring selection of the final plan, we make it possible for the user to trade off running time and result quality.
%\item \textbf{Proof of optimality:}
% We prove that our cost-model and planning algorithm produce a sound and complete final Pareto frontier, making the final plan choice optimal under our search space and the user's preferences.
\end{itemize}

>>>>

A simpler approach might drop all rows with missing values, but not only can this introduce bias into the results, but this may result in discarding all the data.
In contrast to existing systems~\cite{burgette2010multiple,akande2015empirical}, \ProjectName{} avoids imputing over the entire data set, as done with traditional methods. Specifically, \ProjectName{}
carefully plans the placement  imputation / row drop operations inside the query plan, resulting in a 
significant speedup in query execution while minimizing the number of dropped rows.  Unlike previous work, the focus on of our work is not on the imputation algorithms themselves (we can employ almost any such algorithm), but rather on placing imputation operations
optimally in query plans.   Specifically, our optimization algorithm generates the Pareto-optimal frontier of plans that trade imputation cost for result quality.

@@

\subsection{Contributions}
\ProjectName{} leverages a number of contributions to minimize imputation time while producing comparable results to traditional approaches,
increasing productivity for users who frequently encounter missing values in their data. These contributions include:
\begin{itemize}
\item \textbf{Relational algebra extended with imputation:}
  We extend the standard relational algebra with two new operators to represent imputation operations: impute ($\mu$) and drop ($\delta$).
  Impute operation fills in missing data values using an iterative regression model (CE-CART)~\cite{burgette2010multiple}.
  Drop operation simply drops tuples which contain \nullv{} values.
\item \textbf{Model of imputation quality and cost:}
  We extend the traditional cost model for query plans to incorporate a measure of the quality of the imputations performed.
  We use the cost model to abstract over the imputation algorithm used.
  To add an imputation technique, it is sufficient to characterize it with two functions: one to describe its running time and one to describe the quality of its results.
\item \textbf{Query planning with imputation:}
  We present the first query planning algorithm to jointly optimize for running time and the quality of the imputed results.
  It does so by maintaining multiple sets of Pareto optimal plans according to the cost model.
  By deferring selection of the final plan, we make it possible for the user to trade off running time and result quality.
%\item \textbf{Proof of optimality:}
% We prove that our cost-model and planning algorithm produce a sound and complete final Pareto frontier, making the final plan choice optimal under our search space and the user's preferences.
\end{itemize}

### 2017-03-01 paper/relatedwork.tex
paper: polishing changes

In~\cite{akande2015empirical}, Akande et al analyze the performance of various
multiple imputation techniques on the American Community Survey dataset. 
The computational difficulties of imputing on large base
tables are well known and can limit approaches. For example, Akande finds that
one approach (MI-GLM) is prohibitively expensive when attempting to impute on data that
includes variables with potentially large domains (ten categories in their case).
In contrast, \ProjectName{} allows users to specify a tradeoff between
imputation quality and runtime performance, allowing users to perform queries
directly on the entire dataset. Furthermore,
the query planner's imputation is guided by the requirements of each specific
query's operators, rather than requiring broad assumptions about query
workloads.

@@

Database system developers and others have worked on techniques to automatically
detect dirty values, whether missing or otherwise, and rectify the errors if
possible. A survey of methods and systems is provided in
\cite{hellerstein2008quantitative}.

In \cite{wolf2007query}, queries over databases with missing values are
processed in a statistical approach, taking advantage of correlations between
variables. Both tuples that match the original query as well as a ranking of
tuples with incomplete data that may match are returned. Our work differs in
that we allow any well-formed statistical technique to be used for imputation
and focus on returning results to the analyst as if the database had been
complete.

Designers of data stream processing systems frequently confront missing values
and consider them carefully in query processing. Often, if sensor error is the
cause of missing values, values can be imputed with high confidence. In
\cite{fernandez2009inter}, feedback punctuation is used to dynamically
reconfigure the query plan for state-dependent
optimizations as data continue to arrive. One of the applications of this framework is to avoid
expensive imputations as real-time conditions change. 
\ProjectName{} focuses on handling
missing values, rather than providing a general framework for planning feedback. Furthermore,
it provides the ability
to trade imputation quality for runtime performance. 
%We prove our
%algorithm and cost model produces an optimal set of plans to choose from,
%given the constraints of the search space.

BayesDB \cite{mansinghka2015bayesdb} provides users with a simple interface to 
leverage statistical inference techniques in a database. Non-experts
can use a simple declarative language (an extension of SQL), to specify models
which allow missing value imputation, amongst other broader functionality.
Experts can further customize strategies and express domain knowledge to
improve performance and accuracy.

@@

\ProjectName's cost-based query planner 
is partially based on the seminal work developed for System R's query planning\cite{blasgen1981system}.
However, in contrast to System R, \ProjectName{} performs additional optimizations for imputing missing data and
uses histogram transformations to account for the changing nature of missing values.

\subsection{Forecasting and Databases}
Parisi et al\cite{parisi2011embedding} introduce the idea of incorporating time-series forecast operators into
databases, along with the necessary relational algebra extensions. Their work explores the theoretical
properties of forecast operators and generalizes them into a family of operators, distinguished by
the type of predictions returned. They highlight the use of forecasting for replacing missing values.
In their follow on work\cite{parisi2013temporal}, they identify various equivalence and containment
relationships when using forecast operators, which could be used to perform query plan transformations that guarantee the same result. They
explore forecast-first and forecast-last plans, which perform forecasting operations before and after executing all traditional
relational operators, respectively.

@@

Similarly to Fa, \ProjectName{} provides a declarative language, as
a subset of standard SQL. \ProjectName{}, however, is not searching the space of possible
imputation models, but rather the space of query plans that incorporate imputation operators. Another major point
of distinction between \ProjectName{} and Fa is that the latter doesn't consider tradeoffs between accuracy and computation time, but rather returns the most accurate forecast (with some stopping criterion).

>>>>

In~\cite{akande2015empirical}, Akande et al analyze the performance of various
multiple imputation techniques on the American Community Survey dataset. 
The computational difficulties of imputing on large base
tables are well-known and can limit approaches. For example, Akande finds that
one approach (MI-GLM) is prohibitively expensive when attempting to impute on data that
includes variables with potentially large domains (ten categories in their case).
In contrast, \ProjectName{} allows users to specify a tradeoff between
imputation quality and runtime performance, allowing users to perform queries
directly on the entire dataset. Furthermore,
the query planner's imputation is guided by the requirements of each specific
query's operators, rather than requiring broad assumptions about query
workloads.

@@

Database system developers and others have worked on techniques to automatically
detect dirty values, whether missing or otherwise, and rectify the errors if
possible. A survey of methods and systems is provided in~\cite{hellerstein2008quantitative}.

In~\cite{wolf2007query}, queries over databases with missing values are
processed using a statistical approach, taking advantage of correlations between
variables. The tuples that match the original query as well as a ranking of
tuples with incomplete data that may match are returned. Our work differs in
that we allow any well-formed statistical technique to be used for imputation
and focus on returning results to the analyst as if the database had been
complete.

Designers of data stream processing systems frequently confront missing values
and consider them carefully in query processing. Often, if sensor error is the
cause of missing values, values can be imputed with high confidence. In~\cite{fernandez2009inter}, feedback punctuation is used to dynamically
reconfigure the query plan for state-dependent
optimizations as data continue to arrive. One of the applications of this framework is to avoid
expensive imputations as real-time conditions change. 
\ProjectName{} focuses on handling
missing values, rather than providing a general framework for planning feedback. Furthermore,
it provides the ability
to trade imputation quality for runtime performance. 
%We prove our
%algorithm and cost model produces an optimal set of plans to choose from,
%given the constraints of the search space.

BayesDB~\cite{mansinghka2015bayesdb} provides users with a simple interface to 
leverage statistical inference techniques in a database. Non-experts
can use a simple declarative language (an extension of SQL), to specify models
which allow missing value imputation, amongst other broader functionality.
Experts can further customize strategies and express domain knowledge to
improve performance and accuracy.

@@

\ProjectName's cost-based query planner 
is partially based on the seminal work developed for System R's query planning~\cite{blasgen1981system}.
However, in contrast to System R, \ProjectName{} performs additional optimizations for imputing missing data and
uses histogram transformations to account for the changing nature of missing values.

\subsection{Forecasting and Databases}
Parisi et al.~\cite{parisi2011embedding} introduce the idea of incorporating time-series forecast operators into
databases, along with the necessary relational algebra extensions. Their work explores the theoretical
properties of forecast operators and generalizes them into a family of operators, distinguished by
the type of predictions returned. They highlight the use of forecasting for replacing missing values.
In their follow on work~\cite{parisi2013temporal}, they identify various equivalence and containment
relationships when using forecast operators, which could be used to perform query plan transformations that guarantee the same result. They
explore forecast-first and forecast-last plans, which perform forecasting operations before and after executing all traditional
relational operators, respectively.

@@

Similarly to Fa, \ProjectName{} provides a declarative language, as
a subset of standard SQL.\@ \ProjectName{}, however, is not searching the space of possible
imputation models, but rather the space of query plans that incorporate imputation operators. Another major point
of distinction between \ProjectName{} and Fa is that the latter doesn't consider tradeoffs between accuracy and computation time, but rather returns the most accurate forecast (with some stopping criterion).
